---
title: "Examining the Impact of Covid-19 on Penicillin Antibiotic Prescriptions across Scottish Health Boards"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    theme: flatly
date: "2025-11-27"
---
# Introduction  

Antibiotics are essential to treat and decrease the spread of infections within communities. Increasing rates of AMR (Antimicrobial Resistance), has prompted investigation into antibiotic prescribing patterns, revealing inappropriate and overuse of antibiotics (Nowakowska et al., 2019). In Scotland, Antibiotics are most commonly prescribed in GP practices, often for the treatment of respiratory diseases. However, some cases these antibiotics are unnecessary, which drives bacterial resistance to antibiotics (Nowakowska et al., 2019; Malcolm et al., 2020). During the Covid-19 pandemic, lock down restrictions on where the public could go, decreased the number of GP trips, and social mixing in the community subsequently lowering rate of antibiotic prescriptions (Malcolm et al., 2020).  

This analysis will examine changes in Penicillin antibiotic prescriptions across Scottish Health boards, by assessing prescription data from all Scottish Health boards from 2024 (4 years after the pandemic began), 2020 (during the pandemic), and 2016 (4 years before the pandemic), to view how the number and types of penicillin antibiotics prescribed changed as a result of the pandemic. Examining changes in penicillin antibiotic prescribing as a result of Covid-19 could give powerful insights into ways in which antibiotic prescribing could be monitored to reduce the emergence and burden of AMR. 

# Data Processsing

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Loading R packages 

```{r}
library(tidyverse) #For ggplot
library(janitor) #Needed for cleaning names
library(here) #To load data from within a file
library(plotly) #Needed to make the graph interactive
library(gt) #Loading package to make table
library(RColorBrewer) #To make colour palette
```

## Data Loading
### Prescription Data

Listed all urls from data sets into a csv file to load numerous data sets similtaneously. A function was created to load data sets from every month of 2024, 2020, and 2016 from NHS open data website, that cover prescriptions by prescriber location. The data was read and cleaned to make logical column names, then filtered to select only penicillin antibiotic prescriptions in Scotland, using the "cillin" ending they have in common. Data sets were joined and merged to form a full data set of the 3 years prescribing data. Antibiotic prescription names were simplified to show only antibiotic types, regardless of dosage. Months were removed from year labels to focus on annual summaries.
```{r}
#Select the column in csv file called links which contains Urls 
list_urls <- read_csv(here("data","file_links.csv")) %>% 
  select(Links) %>%  
  
#Pull takes Urls from this csv and uploads into a list of r values called list_urls
  pull() 

#Function to read csv files, and renaming columns to have the same name
process_csv <- function(url) {
  prescription_data <- read_csv(url) %>% clean_names() %>% 
 rename_with(.fn = ~ ifelse(.x %in% c("hbt2014", "hbt"), "hb", .x))

#Renaming to logical names
  prescription_data <- prescription_data %>% 
 rename(year = paid_date_month, antibiotic_type = bnf_item_description, prescription_number = paid_quantity) %>% 
    
#Selecting columns to be examined
    select(hb, antibiotic_type, prescription_number, year) %>%
    
#Filtering only antibiotics, removing columns with no prescription data
    filter(str_detect(antibiotic_type, "CILLIN"), !is.na(hb)) %>% 

#Editing prescription and year columns to remove months and dosages
    mutate(antibiotic_type = case_when(str_starts(antibiotic_type, "AMOX") ~ "AMOXICILLIN", str_starts(antibiotic_type, "BENZYL") ~ "BENZYLPENICILLIN", str_starts(antibiotic_type, "PHENO") ~ "PHENOXYMETHYLPENICILLIN", str_starts(antibiotic_type, "PIV") ~ "PIVMECILLINAM", str_starts(antibiotic_type, "PIPE") ~ "PIPERACILLIN", str_starts(antibiotic_type, "AMPI") ~ "AMPICILLIN", str_starts(antibiotic_type, "FLU") ~ "FLUCLOXACILLIN", TRUE ~ antibiotic_type),year = case_when(str_starts(year, "2020") ~ "2020", str_starts(year, "2024") ~ "2024", str_starts(year, "2016") ~ "2016", TRUE ~ as.character(year))) %>%

#Grouping categories   
    group_by(year, antibiotic_type, hb) %>%  
    
#Adding up all antibiotics of the same type, health board, and year to calculate totals    
    summarise(prescription_number = sum(prescription_number, na.rm = TRUE), .groups = "drop")} 

#Running function with all data in list of urls
combined_prescription_data <- map_dfr(list_urls, process_csv)
```
### Health Board Data

```{r}
#Loading health board names data, and cleaning names to change uppercase to lower case with underscores seperating words instead of spaces
health_board_names <- read_csv("https://www.opendata.nhs.scot/dataset/9f942fdb-e59e-44f5-b534-d6e17229cc7b/resource/652ff726-e676-4a20-abda-435b98dd7bdc/download/hb14_hb19.csv") %>% clean_names()
```
### Population Data 

Loading population data from the census 2022, to join with the prescription data to allow calculation of Antibiotic prescriptions per person. Population data from the census needed to be tidied as it is in a format not compatible with R. This involved renaming the empty column as spare, the columns to the same name as the other combined data set. Additionally combining the population of all age and sex groups to calculate total populations of each health board. 
```{r}
#Reading data within the 'data' folder
population <- read_csv(here("data", "healthboardpopulationbyage.csv"), skip = 10) %>%
  
#Putting column names into lower case space seperated by _
  clean_names() %>% 
  
#Renaming columns to same as other data
  rename(Spare = "x6", hb_name = "health_board_area_2019", health_board_population = count) %>% 
  
#Combining all ages and sex to calculate total population of health boards
  filter(age == "All people" & sex == "All people") %>% 
  select(hb_name, health_board_population) %>% 
  
#Putting health board names in same format as health board data
  mutate(hb_name = paste("NHS", hb_name))
```

## Data Joining 

Data must all be joined to allow identification of health boards and comparisons with antibiotic prescription number and population.
```{r}
#Joining function matches the same columns 
all_prescription_and_hb_data <- combined_prescription_data %>% 
  full_join(health_board_names) %>% 
  full_join(population) 
```

## Data Wrangling
### Calculating Antibiotics Per person

Calculating antibiotics per person is necessary since each health board has a different population. In order to accurately compare health board prescriptions.
```{r}
all_prescription_and_hb_data <- all_prescription_and_hb_data %>% 

#Filtering out empty columns 
  filter(!is.na(hb_name)) %>% 
  group_by(antibiotic_type, year, hb_name, health_board_population) %>% 
  summarise(prescription_number = sum(prescription_number, na.rm = TRUE)) %>% 
  
#Calculating antibiotics per person and placing in a new column
  mutate(antibiotics_per_person = prescription_number / health_board_population)
```

### Calculating Differences in Antibiotics Per Person Over the Years

Next the scale at which these differences occurred was examined to determine which health board may had the greatest Covid-19 impact on prescribing practices.
```{r}
ab_differences <- all_prescription_and_hb_data %>%
  group_by(year, hb_name) %>%
  summarise(total_ab_prescriptions = sum(antibiotics_per_person, na.rm = TRUE), .groups = "drop") %>% 
  
#Pivot so each year is a column, id is hb_name
  pivot_wider(names_from = year, values_from = total_ab_prescriptions, id_cols = hb_name) %>% 
  
#Ensure numeric to allow calculations between columns to be conducted 
  mutate(across(c(`2016`, `2020`, `2024`), as.numeric)) %>% 
  
#Remove empty rows
  filter(!is.na(`2016`) & !is.na(`2020`) & !is.na(`2024`)) %>%  
  
#Calculate percentage change for each comparison period
  mutate(pct_change_2020_2016 = ((`2020` - `2016`) / `2016`) * 100,
    pct_change_2024_2020 = ((`2024` - `2020`) / `2020`) * 100) %>%   
  
#Pivot the percentage change columns to long format for plotting
  pivot_longer(cols = starts_with("pct_change"), names_to = "comparison", values_to = "percentage_change") %>%
  
#Changing column names
  mutate(comparison = case_when(
    comparison == "pct_change_2020_2016" ~ "2016 to 2020",
    comparison == "pct_change_2024_2020" ~ "2020 to 2024",
    TRUE ~ comparison)) %>%
  
#Arrange in descending order 
  arrange(desc(percentage_change))
```
## Data Design Functions

### Creating a Custom Theme and Colour Scheme

Creating a theme for graphs, so all have cohesive colours, text sizes, titles, backgrounds. Additionally, adapting a colour palette to use across graphs. 
```{r}
#Defining text sizes, boldness, text angle of graph elements, and background colours
my_theme <- function() {
  list(theme_minimal(base_size = 12),
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8), 
    axis.text.y = element_text(size = 8), 
    axis.title = element_text(face = "bold", size = 10), 
    plot.title = element_text(face = "bold", size = 12, hjust = 0), 
    legend.position = "bottom", 
    legend.title = element_text(face = "bold", size = 10), 
    legend.text = element_text(size = 8), 
    panel.grid.major = element_line(color = "grey85"), 
    panel.grid.minor = element_blank(), 
    strip.text = element_text(face = "bold", size = 8)),
    guides(fill = guide_legend(nrow = 1)))
}
#Adapting a colour palette from a pre made set that has 8 layers, to 14 
my_colours <- colorRampPalette(brewer.pal(8, "Set2"))(14)
```
# Analysis
## Plot 1

Plot 1 examined whether the number of penicillin antibiotics prescribed per person was impacted by the Covid-19 Pandemic, by analysing prescription data from 2016, 2020, and 2024.
```{r}
p1 <- all_prescription_and_hb_data %>% 
  group_by(year, hb_name) %>%
  summarise(total_antibiotics = sum(antibiotics_per_person), .groups = "drop") %>%

#Plotting and adding interactive text to graph 
  ggplot(aes(x = year, y = total_antibiotics, color = hb_name, group = hb_name, text = paste0("Health Board: ", hb_name,"\nYear: ", year,"\nAntibiotics Per Person: ", round(total_antibiotics,2)))) + 
  
#Editing line and point thickness
  geom_line(size = 1) + geom_point(size = 2) + 

#Setting my own colour scheme
  scale_color_manual(values = my_colours) + 

#Adding graph labels 
  labs(x = "Year", y = "Antibiotics Per Person", title = "Change in Antibiotic Prescriptions Over \n Time by Health Board", color = "Health Board") +
  
#Adding graph theme
  my_theme() 

#Adding interactive graph element and adjusting plot size 
ggplotly(p1, tooltip = "text", width = 800, height = 600) 
```
This combined full data set was transformed into a graph, comparing  total number of prescriptions per head in each health board and facet-wrapped across years. It was seen that during Covid-19 in 2020, penicillin antibiotic prescriptions decreased from 2016 before Covid, and increased after Covid-19 (2024). This result is consistent with a study conducted on the impact of Covid 19 on antibiotic prescriptions, that showed a 34% reduction in antibiotic prescriptions for respiratory conditions from 2019 to 2020 (Malcolm et al,. 2020). This decrease could be a result of decreased diagnosed infections due to reduced social interactions and GP visits, lowering antibiotic use (Malcolm et al., 2020). This reduction could have reduced inappropriate penicillin antibiotic prescriptions, which could lower AMR, as studies estimate nearly 69.6% of antibiotic prescriptions to be inappropriate in England, suggesting similar trends in Scotland (Zhong et al., 2023). This data only assessed community antibiotic prescriptions though, so Covid-19 may have impacted hospital penicillin antibiotic prescriptions differently. 

## Plot 2

Next, the proportions of different types of penicillin antibiotics prescribed were examined across health boards to see if Covid-19 impacted these.
```{r}
#Calculating proportions of totals to visualise different amounts of each antibiotic prescribed
p2 <- all_prescription_and_hb_data %>%
  group_by(hb_name, year) %>%
  
#Renaming columns and adding interactive text labels
  mutate(total_antibiotics = sum(antibiotics_per_person), prop_antibiotic = antibiotics_per_person / total_antibiotics) %>% 
  ggplot(aes(x = year, y = prop_antibiotic, fill = antibiotic_type, 
             text = paste0("Health Board: ", hb_name, "\nYear: ", year, "\nAntibiotic: ", antibiotic_type, "\nProportion:", scales::percent(prop_antibiotic, accuracy = 0.1)))) +
  
#Placing bars on top of each other to see proportions better
  geom_col(position = "stack") + 
  
#Faceting by health board 
  facet_wrap(~hb_name) + 
  
#Putting scales in percentage form
  scale_fill_manual(values = my_colours) + 
  scale_y_continuous(labels = scales::percent_format()) + 
  labs(x = "Year", y = "Proportion of Antibiotic Types (%)", 
       title = "Proportions of Antibiotic Types Across Health Boards Over Years", fill = "Antibiotic Type") +
  my_theme() 

ggplotly(p2, tooltip = "text", width = 800, height = 600) 
```
Across all 3 years, Amoxicillin was the most commonly prescribed antibiotic, followed by phenoxymethylpenicillin. Both of these  are broad spectrum antibiotics meaning they are prescribed to treat a range of infections, which could explain why these are most commonly prescribed in Scotland (Nowakowska et al., 2019). Generally the proportion of antibiotic types weren't impacted by Covid-19, with the expection of flucloxacillin, a narrow spectrum antibiotic, which increased in 2024. This is in line with a study by Francis et al, and may be used more due to its specificity, as antibiotic resistance has developed against other commonly prescribed antibiotics (Francis et al., 2016). However, this increase in flucloxacillin should also be treated cautiously, as it represents a selection pressure for resistance to flucloxacillin too (Francis et al., 2016). 

## Plot 3

Next the extent to which antibiotic prescriptions were affected by the pandemic was assessed to see if different health boards were impacted at different scales. Solid line: the percentage change from 2016-2020 (before the pandemic to during the pandemic), Dashed line: percentage change from 2020-2024 (during the pandemic to after the pandemic).
```{r}
p3 <- ab_differences %>%
  
#Defining what hover captions will say
  mutate(hb_name = fct_reorder(hb_name, percentage_change),
         direction = if_else(percentage_change > 0, "Increase", "Decrease"), 
         text = paste0("Health Board: ", hb_name, "\nChange: ", 
                       round(percentage_change, 3), " per person (", direction, ")", "\nPeriod:", comparison)) %>%
  
#Plotting graph and defining different line types for comparisons
  ggplot(aes(x = hb_name, y = percentage_change, color = comparison, group = comparison,
             text = text, linetype = comparison)) +
#Adding a line at 0
  geom_hline(yintercept = 0, color = "gray50", linetype = "dashed") +  
  geom_line(size = 1) +
  geom_point(size = 2) +
  scale_color_manual(values = my_colours) +
  scale_linetype_manual(values = c("solid", "dashed")) + 
  labs(x = "Health Board", y = "Percentage Change in Antibiotics Per Person",
       title = "Percentage Change in Antibiotic Prescriptions: Covid-19 Impact",
       subtitle = "Comparison across three time periods",
       color = "Period",
       linetype = "Period") +  
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  my_theme()

ggplotly(p3, tooltip = "text", width = 1000, height = 600)
```
Here it can be seen that, when the pandemic begun, antibiotic prescriptions decreased across all health boards, increased again after the pandemic. The biggest change in antibiotic prescriptions per person from during 2020 to 2024 was seen in the Western Isles health board and the largest change from 2016 to 2020 was in Shetland. This may have been because restrictions limited movement flow off the islands which may have limited infection spread during the pandemic, but once restrictions lifted there may have been increased travel to and from the mainland, which could have increased the number of infections, and subsequently the need for antibiotics (Malcolm et al., 2020). However, it must be said that some health boards did not supply prescription information for some years so results may not be representitive of actual changes in prescriptions across health boards.

## Table of Results

A table of results shows the total number of each antibiotic prescribed per person across all health boards each year. 
```{r}
#Editing data to produce tibble data 
tibble_data <- all_prescription_and_hb_data %>%
  group_by(antibiotic_type, year) %>% 
  summarise(total_antibiotics = sum(antibiotics_per_person, na.rm = TRUE), .groups = "drop") %>% 
  filter(!is.na(total_antibiotics))

#Producing table, deciding column headings
tibble_data %>%
  gt(groupname_col = "year") %>%
  cols_label(antibiotic_type = "Antibiotic Type", total_antibiotics = "Total Antibiotics Per Person Across All Health Boards") %>%
  tab_header(title = "Covid-19 Impact on Penicillin Antibiotic Prescriptions", subtitle = "Total Data from Scottish health boards 2016, 2020, 2024") %>%
  
#Only showing 4 decimal places of values 
  fmt_number(columns = total_antibiotics, decimals = 4) %>% 
  
#Designing column headers
 tab_style(style = list(cell_text(weight = "bold")),
  locations = list(cells_title(groups = "title"), cells_column_labels(), cells_column_spanners(), cells_row_groups())) %>% 
tab_style(style = cell_fill(color = "white"), locations = cells_row_groups()
) %>% 
data_color(columns = antibiotic_type, method = "factor", palette = my_colours) %>%
  
#Adding tabs to click where the information is from
tab_source_note(source_note = html('📊 <a href="https://www.opendata.nhs.scot/dataset/prescriptions-in-the-community" target="_blank" style="text-decoration:none;"> Click for Prescription Datasets</a>', '📊 <a href= "https://www.scotlandscensus.gov.uk/webapi/jsf/tableView/tableView.xhtml" target="_blank" style="text-decoration:none;"> Click for Census Data </a>')) 
```

# Conclusion

In conclusion, there was a change in penicillin antibiotic prescribing patterns across Scottish health boards during the Covid-19 pandemic. Prescriptions decreased from 2016, and increased again in 2024 following the pandemic. Additionally, amoxicillin was the most common antibiotic prescribed across all years, though flucloxacillin prescriptions increased in 2024, from both 2016 and 2020. Some health boards, such as Shetland and the Western Isles had bigger changes in penicillin antibiotic prescriptions too, which could have been due to different travel lock down restrictions. Reduced prescriptions could have been due to less GP visits, and cases of infection, reducing the number of  prescriptions. Changes in types of antibiotics could also have been an effect of the pandemic, or due to changes in prescription guidelines put in place after the Covid-19 pandemic, due to the threats of AMR. However, more in depth research into other types of antibiotics, and specific lock down restrictions present in each health board could provide more in depth analysis to the impact of Covid-19 on penicillin antibiotic Prescriptions in Scottish Communities. 

# References

```{r}
#Creating a tibble of references and urls 
references <- tibble(References = c("Francis, N.A., Hood, K., Lyons, R. and Butler, C.C., 2016. Understanding flucloxacillin prescribing trends and treatment non-response in UK primary care: a Clinical Practice Research Datalink (CPRD) study. Journal of Antimicrobial Chemotherapy, 71(7), pp.2037-2046.", "Malcolm, W., Seaton, R.A., Haddock, G., Baxter, L., Thirlwell, S., Russell, P., Cooper, L., Thomson, A. and Sneddon, J., 2020. Impact of the COVID-19 pandemic on community antibiotic prescribing in Scotland. JAC-Antimicrobial Resistance, 2(4), p.dlaa105.", "Nowakowska, M., Van Staa, T., Mölter, A., Ashcroft, D.M., Tsang, J.Y., White, A., Welfare, W. and Palin, V., 2019. Antibiotic choice in UK general practice: rates and drivers of potentially inappropriate antibiotic prescribing. Journal of Antimicrobial Chemotherapy, 74(11), pp.3371-3378", "Zhong, X., Pate, A., Yang, Y.T., Fahmi, A., Ashcroft, D.M., Goldacre, B., MacKenna, B., Mehrkar, A., Bacon, S.C., Massey, J. and Fisher, L., 2023. The impact of COVID-19 on antibiotic prescribing in primary care in England: evaluation and risk prediction of appropriateness of type and repeat prescribing. Journal of Infection, 87(1), pp.1-11."), URL = c("https://doi.org/10.1093/jac/dkw084", "https://doi.org/10.1093/jacamr/dlaa105", " https://doi.org/10.1093/jac/dkz345", "https://doi.org/10.1016/j.jinf.2023.05.010")) %>%
  
mutate(Clickable_Reference = paste0('<a href="', URL, '" target="_blank" style="color: #333; text-decoration: none;">', References, '</a>')) 

#Making clickable references
references %>%
  select(Clickable_Reference) %>%
  gt() %>%
  fmt_markdown(columns = Clickable_Reference) %>% 
  cols_label(Clickable_Reference = "References") %>%
  
#Making column heading text bold
  tab_header(title = "Reference Table", subtitle = "Harvard Formatted") %>% 
  tab_style(style = cell_text(weight = "bold"), locations = cells_title())
```
