---
title: "Assessment"
output: html_document
date: "2025-10-31"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Loading R packages
```{r}
library(tidyverse)
library(janitor)
library(here)
```

Loading 2024 data 
```{r}
jan_2024 <- read_csv("https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/d3eaaf84-0f3b-4fb8-9460-e33503095fbe/download/pitc202401.csv") %>% 
  clean_names() %>% 
  filter(str_detect(bnf_item_description, "CILLIN")) %>% 
  select(hbt, bnf_item_description, paid_quantity, paid_date_month) %>% 
  group_by(bnf_item_description, paid_date_month)

feb_2024 <- read_csv("https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/86fda652-0e1d-48bb-9368-aa2a560d925b/download/pitc202402.csv") %>% 
  clean_names() %>% 
  filter(str_detect(bnf_item_description, "CILLIN")) %>% 
  select(hbt, bnf_item_description, paid_quantity, paid_date_month) %>% 
  group_by(bnf_item_description, paid_date_month)

mar_2024 <- read_csv("https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/a42762ac-47cb-4fb6-b9b1-2478a588c0ed/download/pitc202403.csv") %>% 
  clean_names() %>% 
  filter(str_detect(bnf_item_description, "CILLIN")) %>% 
  select(hbt, bnf_item_description, paid_quantity, paid_date_month) %>% 
  group_by(bnf_item_description, paid_date_month)

apr_2024 <- read_csv("https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/409a02f2-b77c-47a0-917d-4a5a1c90f182/download/pitc202404.csv") %>% 
  clean_names() %>% 
  filter(str_detect(bnf_item_description, "CILLIN")) %>% 
  select(hbt, bnf_item_description, paid_quantity, paid_date_month) %>% 
  group_by(bnf_item_description, paid_date_month)

  
may_2024 <- read_csv("https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/5fbbd126-6166-4249-9620-7ed78e877297/download/pitc202405.csv") %>% 
  clean_names() %>% 
  filter(str_detect(bnf_item_description, "CILLIN")) %>% 
  select(hbt, bnf_item_description, paid_quantity, paid_date_month) %>% 
  group_by(bnf_item_description, paid_date_month)

june_2024 <- read_csv("https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/95f7f250-bd04-4e4a-b853-5df75b00a632/download/pitc202406.csv") %>% 
  clean_names() %>% 
  filter(str_detect(bnf_item_description, "CILLIN")) %>% 
  select(hbt, bnf_item_description, paid_quantity, paid_date_month) %>% 
  group_by(bnf_item_description, paid_date_month)


july_2024 <- read_csv("https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/1cb425fc-640c-4b37-9013-f8e97f274085/download/pitc202407.csv") %>% 
  clean_names() %>% 
  filter(str_detect(bnf_item_description, "CILLIN")) %>% 
  select(hbt, bnf_item_description, paid_quantity, paid_date_month) %>% 
  group_by(bnf_item_description, paid_date_month)


aug_2024 <- read_csv("https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/e4985a62-9d59-4e71-8800-3f7ca29ffe0c/download/pitc202408.csv") %>% 
  clean_names() %>% 
  filter(str_detect(bnf_item_description, "CILLIN")) %>% 
  select(hbt, bnf_item_description, paid_quantity, paid_date_month) %>% 
  group_by(bnf_item_description, paid_date_month)


sep_2024 <- read_csv("https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/d1fbede3-98c4-436e-9e75-2ed807a36075/download/pitc202409.csv") %>% 
  clean_names() %>% 
  filter(str_detect(bnf_item_description, "CILLIN")) %>% 
  select(hbt, bnf_item_description, paid_quantity, paid_date_month) %>% 
  group_by(bnf_item_description, paid_date_month)

  
oct_2024 <- read_csv("https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/0d27f6e8-5f8b-42c4-9fe9-c44389af2d3e/download/pitc202410.csv") %>% 
  clean_names() %>% 
  filter(str_detect(bnf_item_description, "CILLIN")) %>% 
  select(hbt, bnf_item_description, paid_quantity, paid_date_month) %>% 
  group_by(bnf_item_description, paid_date_month)

nov_2024 <- read_csv("https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/1b97dffc-310e-4a30-bf9c-22fddff3f499/download/pitc202411.csv") %>% 
  clean_names() %>% 
  filter(str_detect(bnf_item_description, "CILLIN")) %>% 
  select(hbt, bnf_item_description, paid_quantity, paid_date_month) %>% 
  group_by(bnf_item_description, paid_date_month)

 
dec_2024 <- read_csv("https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/e4380152-3d16-446b-9072-108e93cf5caa/download/pitc202412.csv") %>% 
  clean_names() %>% 
  filter(str_detect(bnf_item_description, "CILLIN")) %>% 
  select(hbt, bnf_item_description, paid_quantity, paid_date_month) %>% 
  group_by(bnf_item_description, paid_date_month)


```
Loading 2020 Data 
```{r}
jan_2020 <- read_csv("https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/e5c841f2-3e16-428b-97db-0798ec7a5fb4/download/pitc202001.csv") %>% 
  clean_names() %>% 
  filter(str_detect(bnf_item_description, "CILLIN")) %>% 
  select(hbt, bnf_item_description, paid_quantity, paid_date_month) %>% 
  group_by(bnf_item_description, paid_date_month)

feb_2020 <- read_csv("https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/af121e7c-4124-4955-92e9-a9580ccd469e/download/pitc202002.csv") %>% 
  clean_names() %>% 
  filter(str_detect(bnf_item_description, "CILLIN")) %>% 
  select(hbt, bnf_item_description, paid_quantity, paid_date_month) %>% 
  group_by(bnf_item_description, paid_date_month)

mar_2020 <- read_csv("https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/9581bd8d-5568-4462-93a6-6d7bfbbe7cbc/download/pitc202003.csv") %>% 
  clean_names() %>% 
  filter(str_detect(bnf_item_description, "CILLIN")) %>% 
  select(hbt, bnf_item_description, paid_quantity, paid_date_month) %>% 
  group_by(bnf_item_description, paid_date_month)

apr_2020 <- read_csv("https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/9cdc0526-21ab-43de-b832-bc032cd31b24/download/pitc202004.csv") %>% 
  clean_names() %>% 
  filter(str_detect(bnf_item_description, "CILLIN")) %>% 
  select(hbt, bnf_item_description, paid_quantity, paid_date_month) %>% 
  group_by(bnf_item_description, paid_date_month)

may_2020 <- read_csv("https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/f2ffa7b3-3f93-470a-8125-880afb9aafe0/download/pitc202005.csv") %>% 
  clean_names() %>% 
  filter(str_detect(bnf_item_description, "CILLIN")) %>% 
  select(hbt, bnf_item_description, paid_quantity, paid_date_month) %>% 
  group_by(bnf_item_description, paid_date_month)

june_2020 <- read_csv("https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/64131d22-ab47-43ff-9674-ebe8ed7edbd7/download/pitc202006.csv") %>% 
  clean_names() %>% 
  filter(str_detect(bnf_item_description, "CILLIN")) %>% 
  select(hbt, bnf_item_description, paid_quantity, paid_date_month) %>% 
  group_by(bnf_item_description, paid_date_month)

july_2020 <- read_csv("https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/80b6ddb1-f09b-4d76-9927-da1e118b01ff/download/pitc202007.csv") %>% 
  clean_names() %>% 
  filter(str_detect(bnf_item_description, "CILLIN")) %>% 
  select(hbt, bnf_item_description, paid_quantity, paid_date_month) %>% 
  group_by(bnf_item_description, paid_date_month)

aug_2020 <- read_csv("https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/5fd5f56e-d1d2-4f22-b4f7-224fbd50dca3/download/pitc202008.csv") %>% 
  clean_names() %>% 
  filter(str_detect(bnf_item_description, "CILLIN")) %>% 
  select(hbt, bnf_item_description, paid_quantity, paid_date_month) %>% 
  group_by(bnf_item_description, paid_date_month)

sep_2020 <- read_csv("https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/e1b0cfdd-d184-4ebc-bd93-f1a10d84ead0/download/pitc202009.csv") %>% 
  clean_names() %>% 
  filter(str_detect(bnf_item_description, "CILLIN")) %>% 
  select(hbt, bnf_item_description, paid_quantity, paid_date_month) %>% 
  group_by(bnf_item_description, paid_date_month)

oct_2020 <- read_csv("https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/83bf4892-d246-41f8-a6ca-d44d18e2a2ca/download/pitc202010.csv") %>% 
  clean_names() %>% 
  filter(str_detect(bnf_item_description, "CILLIN")) %>% 
  select(hbt, bnf_item_description, paid_quantity, paid_date_month) %>% 
  group_by(bnf_item_description, paid_date_month)

nov_2020 <- read_csv("https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/ab8aa642-2562-4abb-a360-5c5f9e7e349c/download/pitc202011.csv") %>% 
  clean_names() %>% 
  filter(str_detect(bnf_item_description, "CILLIN")) %>% 
  select(hbt, bnf_item_description, paid_quantity, paid_date_month) %>% 
  group_by(bnf_item_description, paid_date_month)

dec_2020 <- read_csv("https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/0c033702-4d88-4f2d-989c-a709b1f4529e/download/pitc202012.csv") %>% 
  clean_names() %>% 
  filter(str_detect(bnf_item_description, "CILLIN")) %>% 
  select(hbt, bnf_item_description, paid_quantity, paid_date_month) %>% 
  group_by(bnf_item_description, paid_date_month)
```
Loading 2016 data 
```{r}
jan_2016 <- read_csv("https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/7c487db9-f450-4319-8b18-4b825380692b/download/pitc201601.csv") %>% 
  clean_names() %>% 
  filter(str_detect(bnf_item_description, "CILLIN")) %>% 
  select(hbt2014, bnf_item_description, paid_quantity, paid_date_month) %>% 
  group_by(bnf_item_description, paid_date_month)

feb_2016 <- read_csv("https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/0727befa-5530-4c1d-aff0-ccb7554eff7a/download/pitc201602.csv") %>% 
  clean_names() %>% 
  filter(str_detect(bnf_item_description, "CILLIN")) %>% 
  select(hbt2014, bnf_item_description, paid_quantity, paid_date_month) %>% 
  group_by(bnf_item_description, paid_date_month)

mar_2016 <- read_csv("https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/20ad9575-4bd7-4bf4-af41-fd36071e2301/download/pitc201603.csv") %>% 
  clean_names() %>% 
  filter(str_detect(bnf_item_description, "CILLIN")) %>% 
  select(hbt2014, bnf_item_description, paid_quantity, paid_date_month) %>% 
  group_by(bnf_item_description, paid_date_month)

apr_2016 <- read_csv("https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/e44eb789-b8e0-41d9-a264-eba18b2d2473/download/pitc201604.csv") %>% 
  clean_names() %>% 
  filter(str_detect(bnf_item_description, "CILLIN")) %>% 
  select(hbt2014, bnf_item_description, paid_quantity, paid_date_month) %>% 
  group_by(bnf_item_description, paid_date_month)

may_2016 <- read_csv("https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/90ebb5ed-bc22-4bbc-8915-d55ae9d9906e/download/pitc201605.csv") %>% 
  clean_names() %>% 
  filter(str_detect(bnf_item_description, "CILLIN")) %>% 
  select(hbt2014, bnf_item_description, paid_quantity, paid_date_month) %>% 
  group_by(bnf_item_description, paid_date_month)

june_2016 <- read_csv("https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/a636862a-77e0-4c97-ba97-268578534a8e/download/pitc201606.csv") %>% 
  clean_names() %>% 
  filter(str_detect(bnf_item_description, "CILLIN")) %>% 
  select(hbt2014, bnf_item_description, paid_quantity, paid_date_month) %>% 
  group_by(bnf_item_description, paid_date_month)

july_2016 <- read_csv("https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/c12c905b-3a2d-4b7a-b5d2-8c359b786beb/download/pitc201607.csv") %>% 
  clean_names() %>% 
  filter(str_detect(bnf_item_description, "CILLIN")) %>% 
  select(hbt2014, bnf_item_description, paid_quantity, paid_date_month) %>% 
  group_by(bnf_item_description, paid_date_month)

aug_2016 <- read_csv("https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/bbd1d028-b9c6-4a28-b579-eafcaaa44703/download/pitc201608.csv") %>% 
  clean_names() %>% 
  filter(str_detect(bnf_item_description, "CILLIN")) %>% 
  select(hbt2014, bnf_item_description, paid_quantity, paid_date_month) %>% 
  group_by(bnf_item_description, paid_date_month)

sep_2016 <- read_csv("https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/2e0d6f14-3752-4c42-9a02-34d867132737/download/pitc201609.csv") %>% 
  clean_names() %>% 
  filter(str_detect(bnf_item_description, "CILLIN")) %>% 
  select(hbt2014, bnf_item_description, paid_quantity, paid_date_month) %>% 
  group_by(bnf_item_description, paid_date_month)

oct_2016 <- read_csv("https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/f22c20c1-6066-4f7d-a939-3ae654d1467a/download/pitc201610.csv") %>% 
  clean_names() %>% 
  filter(str_detect(bnf_item_description, "CILLIN")) %>% 
  select(hbt2014, hbbnf_item_description, paid_quantity, paid_date_month) %>% 
  group_by(bnf_item_description, paid_date_month)

nov_2016 <- read_csv("https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/5d421ba3-ccba-4d98-9772-e2e042b29a0f/download/pitc201611.csv") %>% 
  clean_names() %>% 
  filter(str_detect(bnf_item_description, "CILLIN")) %>% 
  select(hbt2014, bnf_item_description, paid_quantity, paid_date_month) %>% 
  group_by(bnf_item_description, paid_date_month)

dec_2016 <- read_csv("https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/597ea4bc-f733-45b6-8a60-5d7a6ac4824c/download/pitc201612.csv") %>% 
  clean_names() %>% 
  filter(str_detect(bnf_item_description, "CILLIN")) %>% 
  select(hbt2014, bnf_item_description, paid_quantity, paid_date_month) %>% 
  group_by(bnf_item_description, paid_date_month)
```

Loading Health Board names and population data from the census 
```{r}

health_board_names <- read_csv("https://www.opendata.nhs.scot/dataset/9f942fdb-e59e-44f5-b534-d6e17229cc7b/resource/652ff726-e676-4a20-abda-435b98dd7bdc/download/hb14_hb19.csv") %>% 
  clean_names()

population <- read_csv(here("data", "healthboardpopulationbyage.csv"), skip = 10) %>% 
  clean_names()  

  #maybe need to load previous census population data too

```
Joining months data form each year to produce while year datasets 
```{r}

data_2024 <- list(jan_2024, feb_2024, mar_2024, apr_2024, may_2024, june_2024, july_2024, aug_2024, sep_2024, oct_2024, nov_2024, dec_2024) %>% 
  reduce(full_join) %>% rename(Year = paid_date_month, `Antibiotic Type` = bnf_item_description, `Prescription Number` = paid_quantity) %>% 
  mutate(
    `Antibiotic Type` = case_when(
      str_starts(`Antibiotic Type`, "AMOX") ~ "AMOXICILLIN",
      str_starts(`Antibiotic Type`, "BENZYL") ~ "BENZYLPENICILLIN",
      str_starts(`Antibiotic Type`, "PHENO") ~ "PHENOXYMETHYLPENICILLIN",
      str_starts(`Antibiotic Type`, "PIV") ~ "PIVMECILLINAM",
      str_starts(`Antibiotic Type`, "AMPI") ~ "AMPICILLIN",
      str_starts(`Antibiotic Type`, "FLU") ~ "FLUCLOXACILLIN",
      TRUE ~ `Antibiotic Type`
    ),
    Year = as.character(Year),
    Year = case_when(
      str_starts(Year, "2020") ~ "2020",
      str_starts(Year, "2024") ~ "2024",
      str_starts(Year, "2016") ~ "2016",
      TRUE ~ Year)) 

data_2020 <- list(jan_2020, feb_2020, mar_2024, apr_2020, may_2020, june_2020, july_2020, aug_2020, sep_2020, oct_2020, nov_2020, dec_2020) %>% 
  reduce(full_join) %>% rename(Year = paid_date_month, `Antibiotic Type` = bnf_item_description, `Prescription Number` = paid_quantity) %>% 
  mutate(
    `Antibiotic Type` = case_when(
      str_starts(`Antibiotic Type`, "AMOX") ~ "AMOXICILLIN",
      str_starts(`Antibiotic Type`, "BENZYL") ~ "BENZYLPENICILLIN",
      str_starts(`Antibiotic Type`, "PHENO") ~ "PHENOXYMETHYLPENICILLIN",
      str_starts(`Antibiotic Type`, "PIV") ~ "PIVMECILLINAM",
      str_starts(`Antibiotic Type`, "AMPI") ~ "AMPICILLIN",
      str_starts(`Antibiotic Type`, "FLU") ~ "FLUCLOXACILLIN",
      TRUE ~ `Antibiotic Type`
    ),
    Year = as.character(Year),
    Year = case_when(
      str_starts(Year, "2020") ~ "2020",
      str_starts(Year, "2024") ~ "2024",
      str_starts(Year, "2016") ~ "2016",
      TRUE ~ Year)) 


data_2016 <- list(jan_2016, feb_2016, mar_2016, apr_2016, may_2016, june_2016, july_2016, aug_2016, sep_2016, oct_2016, nov_2016, dec_2016) %>% 
  reduce(full_join) %>% rename(Year = paid_date_month, `Antibiotic Type` = bnf_item_description, `Prescription Number` = paid_quantity) %>% 
  mutate(
    `Antibiotic Type` = case_when(
      str_starts(`Antibiotic Type`, "AMOX") ~ "AMOXICILLIN",
      str_starts(`Antibiotic Type`, "BENZYL") ~ "BENZYLPENICILLIN",
      str_starts(`Antibiotic Type`, "PHENO") ~ "PHENOXYMETHYLPENICILLIN",
      str_starts(`Antibiotic Type`, "PIV") ~ "PIVMECILLINAM",
      str_starts(`Antibiotic Type`, "AMPI") ~ "AMPICILLIN",
      str_starts(`Antibiotic Type`, "FLU") ~ "FLUCLOXACILLIN",
      TRUE ~ `Antibiotic Type`
    ),
    Year = as.character(Year),
    Year = case_when(
      str_starts(Year, "2020") ~ "2020",
      str_starts(Year, "2024") ~ "2024",
      str_starts(Year, "2016") ~ "2016",
      TRUE ~ Year)) 


```


Joining prescription Dataset from all years, with health boards names. 
```{r}
data_2024 <- data_2024 %>%
  left_join(health_board_names, by = join_by(hbt == hb)) 

data_2020 <- data_2020 %>%
  left_join(health_board_names, by = join_by(hbt == hb)) 

data_2016 <- data_2016 %>%
  left_join(health_board_names, by = join_by(hbt2014 == hb)) 

```
Selecting and grouping columns to be examined *need to somehow fix this *
```{r}
data_2024 <- data_2024 %>% 
  select(hb_name, `Antibiotic Type`, `Prescription Number`, Year) %>% 
  filter(!is.na(hb_name)) %>% 
  group_by(`Antibiotic Type`, hb_name, Year) %>% 
  summarise(`Prescription Number` = sum(`Prescription Number`, na.rm = TRUE)) %>% 
  rename('Health Board' = hb_name)

data_2020 <- data_2020 %>% 
  select(hb_name, `Antibiotic Type`, `Prescription Number`, Year) %>% 
  filter(!is.na(hb_name)) %>% 
  group_by(`Antibiotic Type`, hb_name, Year) %>% 
  summarise(`Prescription Number` = sum(`Prescription Number`, na.rm = TRUE)) %>% 
  rename('Health Board' = hb_name)

data_2016 <- data_2016 %>% 
  select(hb_name, `Antibiotic Type`, `Prescription Number`, Year) %>% 
  filter(!is.na(hb_name)) %>% 
  group_by(`Antibiotic Type`, hb_name, Year) %>% 
  summarise(`Prescription Number` = sum(`Prescription Number`, na.rm = TRUE)) %>% 
  rename('Health Board' = hb_name)

```

Wrangling with population data to compare with prescription health board data
```{r}
population <- population %>% 
rename(Spare = "x6",
         hb_name = "health_board_area_2019",
         'Health Board Population' = count) %>% 
  filter(age == "All people" & sex == "All people") %>% 
         select(hb_name, 'Health Board Population' ) %>%
        mutate(hb_name = paste("NHS", hb_name)) 
  
```

Join this data together

```{r}
all_ab_data <- list(data_2016, data_2020, data_2024) %>% 
  reduce(full_join) 
```


Graphing ab type per healthboard per year, shows total ab prescriptions - should probs do this by person though
```{r}
library(dplyr)
library(stringr)

all_ab_data %>% 
  ggplot(aes(x = `Health Board`, y = `Prescription Number`, fill = `Antibiotic Type`)) +
  geom_col(show.legend = TRUE) +
  labs(x = "Health Board Name", 
    y = "Antibiotic Quantity Per Health Board", 
    title = "Different in Number of Antibiotics Prescribed per Health Board, Before, During and After Covid 19 Pandemic", 
    fill = "Antibiotic Type") +
  facet_wrap(~Year) +
  theme_minimal(base_size = 14) +  # Cleaner, more modern look
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12), # Tilt x-axis labels for readability
    axis.text.y = element_text(size = 12),
    axis.title = element_text(face = "bold", size = 14),
    plot.title = element_text(face = "bold", size = 16, hjust = 0),
    legend.position = "bottom",  # Move legend below plot
    legend.title = element_text(face = "bold", size = 13),
    legend.text = element_text(size = 12),
    panel.grid.major = element_line(color = "grey85"),
    panel.grid.minor = element_blank(),
    strip.text = element_text(face = "bold", size = 13)) +
  guides(fill = guide_legend(nrow = 1))
```

Graphing ab types per year- measure most common tyoes rather than per perosn amounts 
```{r}
library(ggplot2)
library(dplyr)

all_ab_data %>% 
  ggplot(aes(x = `Antibiotic Type`, y = `Prescription Number`, fill = `Antibiotic Type`)) +
  geom_col(show.legend = TRUE) +
  labs(
    x = "Antibiotic Type", 
    y = "Antibiotic Quantity Per Health Board", 
    title = "Most Common Antibiotics Prescribed Before, During and After Covid 19 Pandemic", 
    fill = "Health Board"  # Corrected legend label to match fill aesthetic
  ) +
  facet_wrap(~ factor(Year)) +  # Convert Year to factor for safe faceting
  theme_minimal(base_size = 14) +  # Cleaner, modern look
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 12),  # Tilt x labels
    axis.text.y = element_text(size = 12),
    axis.title = element_text(face = "bold", size = 14),
    plot.title = element_text(face = "bold", size = 16, hjust = 0),
    legend.position = "bottom",  # Legend below plot
    legend.title = element_text(face = "bold", size = 13),
    legend.text = element_text(size = 12),
    panel.grid.major = element_line(color = "grey85"),
    panel.grid.minor = element_blank(),
    strip.text = element_text(face = "bold", size = 13)
  ) +
  guides(fill = guide_legend(nrow = 1))

```

Joining Population with prescription data, and calculating ab per person
```{r}
all_ab_data_pop <-  all_ab_data %>%
  left_join(population, by = join_by('Health Board' == hb_name)) %>%
  mutate(`Antibiotics Per Person` = `Prescription Number` / `Health Board Population`)
```

 ab per person administered across different years per each healthboard *try and order bars from increasing to decreasing*
```{r}
all_ab_data_pop %>% 
  ggplot(aes(x = `Health Board`, y = `Antibiotics Per Person`, fill = `Antibiotic Type`)) +
  geom_col(show.legend = TRUE) +
  labs(x = "Health Board Name", 
    y = "Antibiotic Quantity Per Person", 
    title = "Different in Number of Antibiotics Prescribed per person, Before, During and After Covid 19 Pandemic", 
    fill = "Antibiotic Type") +
  facet_wrap(~Year) +
  theme_minimal(base_size = 14) +  # Cleaner, more modern look
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12), # Tilt x-axis labels for readability
    axis.text.y = element_text(size = 12),
    axis.title = element_text(face = "bold", size = 14),
    plot.title = element_text(face = "bold", size = 16, hjust = 0),
    legend.position = "bottom",  # Move legend below plot
    legend.title = element_text(face = "bold", size = 13),
    legend.text = element_text(size = 12),
    panel.grid.major = element_line(color = "grey85"),
    panel.grid.minor = element_blank(),
    strip.text = element_text(face = "bold", size = 13)) +
  guides(fill = guide_legend(nrow = 1))

```

PLot 2 but looking at which hb has biggest difference between years in prescribing *maybe work out difference across years and graph this ? as a map or diff type of graph maybe *
```{r}
all_ab_data_pop %>% 
  ggplot(aes(x = Year, y = `Antibiotics Per Person`, fill = `Antibiotic Type`)) +
  geom_col(show.legend = TRUE) +
  labs(
    x = "Year", 
    y = "Antibiotic Quantity Per Person", 
    title = "Different in Number of Antibiotics Prescribed per person, Before, During and After Covid 19 Pandemic", 
    fill = "Antibiotic Type") +
  facet_wrap(~`Health Board`) +
  theme_minimal(base_size = 14) +  # Cleaner, more modern look
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12), # Tilt x-axis labels for readability
    axis.text.y = element_text(size = 12),
    axis.title = element_text(face = "bold", size = 14),
    plot.title = element_text(face = "bold", size = 16, hjust = 0),
    legend.position = "bottom",  # Move legend below plot
    legend.title = element_text(face = "bold", size = 13),
    legend.text = element_text(size = 12),
    panel.grid.major = element_line(color = "grey85"),
    panel.grid.minor = element_blank(),
    strip.text = element_text(face = "bold", size = 13)) +
  guides(fill = guide_legend(nrow = 1))
```


Table containing data about antibiotic type, total number, number per person, across the 3 years and different healthboards *need to ask whether i need to use old census data for 2016 and 2020*, also need to ask if i should look at more months or just these months 

```{r}

```

