---
title: "Examining the differences in Antibiotic Prescriptions across Scottish health boards, before (2016), During (2020) and after (2024) Covid-19"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    theme: flatly
date: "2025-11-27"
---
# Introduction  
Antibiotics are essential to treat and decrease the spread of infections within communities. Increasing rates of AMR (Antimicrobial Resistance), has prompted investigation into antibiotic prescribing patterns,revealing inappropriate and overuse of antibiotics. In Scotland, Antibiotics are most commonly prescribed in GP practices, often for the treatment of respiratory diseases. However, some cases these antibiotics are unnecessary, which drives bacterial resistance to antibiotics (Malcom et al., 2020). During the Covid-19 pandemic, lock down restrictions on where the public could go, decreased the number of GP trips, and social mixing in the community subsequently lowering rate of antibiotic prescriptions (Malcom et al., 2020).  

This analysis will examine changes in antibiotic prescriptions across Scottish Health boards, by assessing prescription data from all Scottish Health boards from 2024 (4 years after the pandemic began), 2020 (during the pandemic), and 2016 (4 years before the pandemic), to view how the number and types of antibiotics prescribed changed as a result of the pandemic. Examining changes in antibiotic prescribing could give powerful insights into ways in which antibiotic prescribing could be monitored to reduce the emergence and spread of AMR. 

# Data Processsing
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```
## Loading R packages 
```{r}
#packages required for analysing prescription data
library(tidyverse) #for ggplot
library(janitor) #
library(here) #to load data from within a file
library(plotly) #needed to make the graph interactive
library(gt) #loading package to make table
library(RColorBrewer) #to make colour palette
library(htmltools) #to add l
```
## Data Loading

### Prescription Data
Listed all urls from data sets into a csv file to load numerous data sets similtaneously. A function was created to load data sets from every month of 2024, 2020, and 2016 from NHS open data website, that cover prescriptions by prescriber location. The data was read and cleaned to make logical column names, the filtered to select only antibiotic prescriptions in Scotland. Data sets were joined and merged to form a full data set of the 3 years prescribing data. Antibiotic prescription names were simplified to show only antibiotic types, regardless of dosage. Months were removed from year labels to focus on annual summaries.
```{r}
list_urls <- read_csv(here("data","file_links.csv")) %>% 
  select(Links) %>% #here, select is used to select the column in csv file called links which contains urls 
  pull() # pull takes urls from this csv and uploads into a list of r values called list_urls

process_csv <- function(url) {
  prescription_data <- read_csv(url) %>% clean_names() %>% 
 rename_with(.fn = ~ ifelse(.x %in% c("hbt2014", "hbt"), "hb", .x))

prescription_data <- prescription_data %>% 
 rename(year = paid_date_month, antibiotic_type = bnf_item_description, prescription_number = paid_quantity) %>% #renaming to logical names
    select(hb, antibiotic_type, prescription_number, year) %>% #selecting columns to be examined
    filter(str_detect(antibiotic_type, "CILLIN"), !is.na(hb)) %>% #filtering only antibiotics, removing columns with no prescription data
    mutate(antibiotic_type = case_when(str_starts(antibiotic_type, "AMOX") ~ "AMOXICILLIN", str_starts(antibiotic_type, "BENZYL") ~ "BENZYLPENICILLIN", str_starts(antibiotic_type, "PHENO") ~ "PHENOXYMETHYLPENICILLIN", str_starts(antibiotic_type, "PIV") ~ "PIVMECILLINAM", str_starts(antibiotic_type, "PIPE") ~ "PIPERACILLIN", str_starts(antibiotic_type, "AMPI") ~ "AMPICILLIN", str_starts(antibiotic_type, "FLU") ~ "FLUCLOXACILLIN", TRUE ~ antibiotic_type),year = case_when(str_starts(year, "2020") ~ "2020", str_starts(year, "2024") ~ "2024", str_starts(year, "2016") ~ "2016", TRUE ~ as.character(year))) %>% # Editing prescription and year columns to remove months and dosages
    group_by(year, antibiotic_type, hb) %>%  #grouping categories
    summarise(prescription_number = sum(prescription_number, na.rm = TRUE), .groups = "drop")} #adding up all antibiotics of the same type, health board, and year to calculate totals

#running function with all data in list of urls
combined_prescription_data <- map_dfr(list_urls, process_csv)
```
### Health Board 
```{r}
#loading health board names data, and cleaning names to change uppercase to lower case with underscores seperating words instead of spaces
health_board_names <- read_csv("https://www.opendata.nhs.scot/dataset/9f942fdb-e59e-44f5-b534-d6e17229cc7b/resource/652ff726-e676-4a20-abda-435b98dd7bdc/download/hb14_hb19.csv") %>% clean_names()
```
### Population Data 
Loading population data from the census 2022, to join with the prescription data to allow calculation of Antibiotic prescriptions per person. Population data from the census needed to be tidied as it is in a format not compatible with R. This involved renaming the empty column as spare, renaming the columns to the same name as the other combined data set. Additionally combining the population of all age and sex groups to calculate total populations of each health board. 
```{r}
population <- read_csv(here("data", "healthboardpopulationbyage.csv"), skip = 10) %>% #reading data within the 'data' folder
  clean_names() %>% #putting column names into lower case space seperated by _
  rename(Spare = "x6", hb_name = "health_board_area_2019", health_board_population = count) %>% #renaming columns to same as other data
  filter(age == "All people" & sex == "All people") %>% #combining all ages and sex to calculate total population of health boards
  select(hb_name, health_board_population) %>% #selecting columns to be examined
  mutate(hb_name = paste("NHS", hb_name))  #putting health board names in same format at health board data
```
## Data joining 
Data must all be joined to allow identification of health boards and comparisons with antibiotic prescription number and population.
```{r}
all_prescription_and_hb_data <- combined_prescription_data %>% 
  full_join(health_board_names) %>% 
  full_join(population) #joining function, matches same columns 
```
## Data Wrangling
### Calculating Antibiotics Per person
```{r}
all_prescription_and_hb_data <- all_prescription_and_hb_data %>% 
  filter(!is.na(hb_name)) %>% #filtering out columns in which healthboards havent supplied appropriate information, i.e. contain no poopulation or prescription datapoints examine
  group_by(antibiotic_type, year, hb_name, health_board_population) %>% 
  summarise(prescription_number = sum(prescription_number, na.rm = TRUE)) %>% #adding up all the prescriptions to get total number of each type in each healthboard
  mutate(antibiotics_per_person = prescription_number / health_board_population) %>%   
  ungroup() %>%  # Important: ungroup before arranging
  arrange(year, antibiotic_type, desc(antibiotics_per_person))  # Sort by year, then antibiotic, then value, group and sort again after to keep ab types together for for better table layout
```
### Calculating Differences in Antibiotics per person over the years
Since it could be seen that the number of antibiotics administered per person did change with Covid-19, next the scale at which these differences occurred was examined to determine which health board may had the greatest Covid impact on prescribing practices
```{r}
# Aggregate first, then pivot
ab_differences <- all_prescription_and_hb_data %>%
  group_by(year, hb_name, antibiotic_type) %>%
  summarise(antibiotics_per_person = sum(antibiotics_per_person, na.rm = TRUE), .groups = "drop") %>%  # Sum up any duplicates
  pivot_wider(names_from = year, values_from = antibiotics_per_person, id_cols = c(hb_name, antibiotic_type)) %>%
  mutate(across(c(`2016`, `2020`, `2024`), ~ as.numeric(.))) %>% # Ensure numeric to allow calculations between columns to be conducted 
  filter(!is.na(`2016`) & !is.na(`2020`) & !is.na(`2024`)) %>% #removing columns with no data, these may have arisen due to not aving data from all 3 years
  mutate(pct_change_2024_2020 = (((`2024` - `2020`)/ `2020`)*100), pct_change_2016_2020 = (((`2016` - `2020`)/ `2020`)*100)) %>% #produces new columns with difference values 
  pivot_longer(cols = starts_with("pct"), names_to = "comparison", values_to = "percentage_change") %>% # Select difference columns, column name for the comparison type
  mutate(comparison = case_when(comparison == "pct_change_2024_2020" ~ "2024 vs 2020", comparison == "pct_change_2016_2020" ~ "2016 vs 2020", TRUE ~ comparison)) %>% 
  arrange(desc(percentage_change)) #arrange from biggest to smallest difference for initial examination of which health board had the greatest change in prescription numbers per person across years.
```
## Data Design Functions
### Creating custom theme and colour scheme
```{r}
#creating my own theme for graphs, so all have cohesive colours, text sizes, titles, backgrounds etc. This saves space within my code and prevents me having to type the same code over again for each graph. 
my_theme <- function() {
  list(theme_minimal(base_size = 12),
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8), axis.text.y = element_text(size = 8), axis.title = element_text(face = "bold", size = 10), plot.title = element_text(face = "bold", size = 12, hjust = 0), legend.position = "bottom", legend.title = element_text(face = "bold", size = 10), legend.text = element_text(size = 8), panel.grid.major = element_line(color = "grey85"), panel.grid.minor = element_blank(), strip.text = element_text(face = "bold", size = 8)),guides(fill = guide_legend(nrow = 1)))
} #defining text sizes of graph elements, and boldness, as well as background colours, angle positioning of text 

#adapting a colour palette from a pre made set that has 8 layers, so it can be used to colour all different health boards 
my_colours <- colorRampPalette(brewer.pal(8, "Set2"))(14)
```
# Analysis
## Plot 1
```{r}
p1 <- all_prescription_and_hb_data %>% 
  group_by(year, hb_name) %>%
  summarise(total_antibiotics = sum(antibiotics_per_person), .groups = "drop") %>% #summarising total antibiotics to see differences across years and hb regradless of type
  ggplot(aes(x = year, y = total_antibiotics, color = hb_name, group = hb_name, text = paste0("Health Board: ", hb_name,"\nYear: ", year,"\nAntibiotics Per Person: ", round(total_antibiotics,2)))) + #adding interactive text to graph
  geom_line(size = 1) + geom_point(size = 2) + #editing line and point thickness
  scale_color_manual(values = my_colours) + #setting my own colour scheme
  labs(x = "Year", y = "Antibiotics Per Person", title = "Change in Antibiotic Prescriptions Over \n Time by Health Board", color = "Health Board") +
  my_theme() #adding my own graph theme

ggplotly(p1, tooltip = "text", width = 800, height = 600) #confirming interactive element of graph
```
This combined full data set was transformed into a graph, comparing  total number of prescriptions per head in each health board and facet-wrapped across years.It was seen that during Covid-19 in 2020, antibiotic prescriptions decreased from 2016 before Covid, and increased after Covid-19 (2024). This result is consistant with a study conducted on the impact of Covid 19 on antibiotic prescriptions, that showed a 34% reduction in antibiotic prescriptions for respiratiory conditions from 2019 to 2020 (Malcom et al,. 2020). This decrease could be a result of decreased diagnosed infections due to reduced social interactions and GP visits, lowering antibiotic use (Malcom et al., 2020). This reduction could have reduced  inappropriate antibiotic prescriptions, which could lower AMR, as studies estimate nearly 69.6% of antibiotic prescriptions to be inappropriate in England, suggesting similar trends in Scotland (Zhong et al., 2023).
## Plot 2
```{r}
p2 <- all_prescription_and_hb_data %>%
  group_by(hb_name, year) %>%
  mutate(total_antibiotics = sum(antibiotics_per_person), #caluclating proportions of totals to visualise different amounts of each antibiotic prescribed
         prop_antibiotic = antibiotics_per_person / total_antibiotics, text = paste0("Health Board: ", hb_name, "\nYear: ", year, "\nAntibiotic: ", antibiotic_type, "\nProportion:", scales::percent(prop_antibiotic, accuracy = 0.1))) %>% ggplot(aes(x = year, y = prop_antibiotic, fill = antibiotic_type, text = paste0("Antibiotic: ", antibiotic_type, "\nYear: ", year, "\nProportion: ", scales::percent(prop_antibiotic)))) + #adding interactive text to graph 
  geom_col(position = "stack") + #placing bars on top of each other to see proportions better
  facet_wrap(~hb_name) + #faceting by health board 
  scale_fill_manual(values = my_colours) + #setting my own colour scheme
  scale_y_continuous(labels = scales::percent_format()) + #making sure scales are in percentage form
  labs(x = "Year", y = "Proportion of Antibiotic Types (%)", title = "Proportions of Antibiotic Types Across Health Boards Over Years", fill = "Antibiotic Type") +
  my_theme() #adding my own graph theme

ggplotly(p2, tooltip = "text", width = 800, height = 600) #confirming interactive element of graph
```
Next a graph was produced showing the most common antibiotic types per year, revealing across all 3 years, Amoxicillin was the most commonly prescribed antibiotic, followed by phenoxymethylpenicillin. Both of these  are broad spectrum antibiotics meaning they are prescribed to treat a range of infections, which could explain why these are most commonly prescribed in Scotland. Additionally it can be seen that the proportion of flucloxacillin, a narrow spectrum antibiotic, increased in 2024. Prescriptions of this may have increased in 2024, due to its specificity, as antibiotic resistance has developed against other commonly prescribed antibiotics (Francis et al., 2016). Though, Pivemicillinam was not commonly prescribed, potentially because of the decreased infection coverage it provides due to its narrow spectrum nature.
## Plot 3
```{r}
#maybe need to reevaluate what this graph is doing 
#plot of ab prescription differences, again using them theme and colour function here
p3 <- ab_differences %>%
  mutate(hb_name = fct_reorder(hb_name, percentage_change), #ordering in increasing order
         text = paste0("Health Board: ", hb_name, "\nChange: ", round(percentage_change, 3), " per person", "\nPeriod: ", comparison)) %>% #adding interactive elements, so that when you hover over graph, values are revealed
  ggplot(aes(x = hb_name, y = percentage_change, color = comparison, group = comparison, text = text)) + geom_line(size = 1) + geom_point(size = 2) + #editing line and point thickness
  labs(x = "Health Board", y = "Percentage Change in Antibiotics Per Person", title = "Change in Antibiotic Prescriptions: COVID-19 Impact", subtitle = "Comparison across three time periods", color = "Period") + 
  scale_color_manual(values = my_colours) + #setting colours to be my own colour scheme 
  my_theme() #adding my own graph theme

ggplotly(p3, tooltip = "text", width = 800, height = 600) #confirming interactive element of graph
```
## Table of Results
```{r}
#editing data to produce tibble data 
tibble_data <- all_prescription_and_hb_data %>%
  select(hb_name, year, antibiotic_type, antibiotics_per_person) %>% 
  group_by(antibiotic_type, year) %>% 
  summarise(total_antibiotics = sum(antibiotics_per_person, na.rm = TRUE), .groups = "drop") %>% 
  filter(!is.na(total_antibiotics))

tibble_data %>%
  gt(groupname_col = "year") %>%
  cols_label(antibiotic_type = "Antibiotic Type", total_antibiotics = "Total Antibiotics Per Person Across All Health Boards") %>%
  tab_header(title = "Antibiotic Prescriptions Before, During and After Covid-19", subtitle = "Total Data from Scottish health boards 2016, 2020, 2024") %>%
  fmt_number(columns = total_antibiotics, decimals = 4) %>%
 tab_style(style = list(cell_text(weight = "bold")),
  locations = list(cells_title(groups = "title"), cells_column_labels(), cells_column_spanners(), cells_row_groups())) %>%
tab_style(style = cell_fill(color = "white"), locations = cells_row_groups()
) %>% 
data_color(columns = antibiotic_type, method = "factor", palette = my_colours) %>%
tab_source_note(source_note = html('📊 <a href="https://www.opendata.nhs.scot/dataset/prescriptions-in-the-community" target="_blank" style="text-decoration:none;">Prescription Datasets</a>', '📊 <a href= "https://www.scotlandscensus.gov.uk/webapi/jsf/tableView/tableView.xhtml" target="_blank" style="text-decoration:none;">Census Data </a>'))
```
# Conclusion
In conclusion, there was a change in antibiotic prescribing patterns across Scottish healthboards during the Covid-19 pandemic. Prescriptions decreased from 2016, and increased again in 2024 following the pandemic. Additionally,  amoxicillin was the most common antibiotic prescribed across all years, but flucloxicillin and pivemicillan prescriptions increased in 2024, from both 2016 and 2020. Reduced prescriptions could have been due to less GP visits, and cases of infection, reducing the number of  prescriptions. Changes in types of antibiotics could also have been an effect of the pandemic, or due to changes in prescription guidelines put in place after the Covid 19 pandemic.
# References
```{r}
references <- tibble(References = c("Francis, N.A., Hood, K., Lyons, R. and Butler, C.C., 2016. Understanding flucloxacillin prescribing trends and treatment non-response in UK primary care: a Clinical Practice Research Datalink (CPRD) study. Journal of Antimicrobial Chemotherapy, 71(7), pp.2037-2046.", "Malcolm, W., Seaton, R.A., Haddock, G., Baxter, L., Thirlwell, S., Russell, P., Cooper, L., Thomson, A. and Sneddon, J., 2020. Impact of the COVID-19 pandemic on community antibiotic prescribing in Scotland. JAC-Antimicrobial Resistance, 2(4), p.dlaa105.", "Zhong, X., Pate, A., Yang, Y.T., Fahmi, A., Ashcroft, D.M., Goldacre, B., MacKenna, B., Mehrkar, A., Bacon, S.C., Massey, J. and Fisher, L., 2023. The impact of COVID-19 on antibiotic prescribing in primary care in England: evaluation and risk prediction of appropriateness of type and repeat prescribing. Journal of Infection, 87(1), pp.1-11."), URL = c("https://doi.org/10.1093/jac/dkw084", "https://doi.org/10.1093/jacamr/dlaa105", "https://doi.org/10.1016/j.jinf.2023.05.010")) %>%
  mutate(Clickable_Reference = paste0('<a href="', URL, '" target="_blank" style="color: #333; text-decoration: none;">', References, '</a>')) #creating a tibble of references and urls to create interactive reference table
references %>%
  select(Clickable_Reference) %>%
  gt() %>%
  fmt_markdown(columns = Clickable_Reference) %>% #making reference table interactive to allow sources to be seen easily
  cols_label(Clickable_Reference = "References") %>%
  tab_header(title = "Reference Table", subtitle = "Harvard Formatted") %>% 
  tab_style(style = cell_text(weight = "bold"), locations = cells_title()) #making column heading text bold
```
